\section{Constant and variables}


Within a program, each declared symbol is associated with a name, also called an
identifier. Identifiers must follow specific rules and are case-sensitive,
meaning that uppercase and lowercase letters are considered different. An
identifier can include letters, numbers, and underscores, but it must contain at
least one letter. Any numbers used in the identifier must be preceded by at
least one letter. An identifier may also begin with any number of underscores.


For example, the identifier \token{__id1__} is valid, whereas
\token{__1id__} is invalid because it starts with a number immediately after the
underscores.

\mynotebox{
  The formal grammar of a identifier is the following: \token{(\_)* [a-zA-Z]([a-zA-Z]|[0-9]|\_)*}
}

Some names that follow the rules for identifiers cannot be used as identifiers
because they are reserved by the language to define specific statements or
operations. These reserved identifiers are called \textit{keywords}. The
complete list of \textit{keywords} is provided below.

\token{alias}, \token{assert}, \token{atomic}, \token{break}, \token{cast},
\token{catch}, \token{class}, \token{const}, \token{copy}, \token{cte},
\token{dcopy}, \token{def}, \token{dg}, \token{dmut}, \token{do}, \token{else},
\token{enum}, \token{expand}, \token{extern}, \token{false}, \token{for},
\token{fn}, \token{future}, \token{if}, \token{impl}, \token{in}, \token{is},
\token{lazy}, \token{let}, \token{loop}, \token{macro}, \token{match},
\token{mod}, \token{move}, \token{mut}, \token{none}, \token{null}, \token{of},
\token{panic}, \token{\_\_pragma}, \token{record}, \token{union}, \token{ref},
\token{return}, \token{spawn}, \token{static}, \token{entity}, \token{throw},
\token{throws}, \token{trait}, \token{true}, \token{typeof}, \token{\_\_test},
\token{unsafe}, \token{use}, \token{\_\_version}, \token{while}, \token{with}.


The meaning and usage of each keyword will be explained throughout
the course of this book.
