\chapter{Fundamentals}
\nopagecolor{}
\label{chap:chap1}
\vspace{-50pt}%

This chapters presents the fundamental characteristic of the Ymir programming language. It introduces the first necessary steps to build a Ymir program, and is also a first step into the basics of low level programming via a high level programming language. In this chapter is also presented the basics on how this book is structured and will guide the reader into fully understanding the concepts presented in this book.

\minitoc%

The developement of a program using the Ymir language follows three fundamental
steps. First, a text editor is used to create text files named source files, in
which source code following the Ymir grammar and semantic is written. These
source files are passed to the compiler that transform source code information
into object files that is made up of machine code. These object files are
finally linked by a linker to form a an executable file. The second and third
step are generally made automatically when invoking the compiler.

\input{chapters/chapter1/figures/compilation_chain}
\vfill%
\pagebreak%
          
A source file is a text file encoded in \textit{utf-8} with the
extension \textit{.yr}. It is important to use the correct extension for the
source file, as the compiler will use it to determine in which language the
source code is written. Indeed, the Ymir language was made to be interoperable
with the C language, and therefore the compiler is also able to compile C source
files, which are using the \textit{.c} extension.

If the source code does not respect the syntax or semantic of the Ymir language,
the compiler uppon invokation, will issue error messages. These messages are
describing the origin of the errors, and are made to be readable by the
programmer to help them correct the problems.
\vfill%
      
\input{chapters/chapter1/section1}
\input{chapters/chapter1/section2}
\input{chapters/chapter1/section3}
\input{chapters/chapter1/section4}
\input{chapters/chapter1/section5}
